Codes for exploratory analysis

### 2025-04-03 code/analysis-code/statistical-analysis.qmd
reg summary
part #4

## Quarto

@@

```{r}
# Load necessary libraries
library(dplyr)

# Define the dependent variable
dependent_var <- "Current_health_expenditure_per_capita_current_US"

# Select 50 variables that are likely to predict the dependent variable
selected_vars <- c(
  # Economic Factors
  "GDP_per_capita_current_US", "GNI_per_capita_Atlas_method_current_US", "GDP_growth_annual_",
  "Inflation_GDP_deflator_annual_", "Domestic_credit_provided_by_financial_sector__of_GDP",
  "Gross_capital_formation__of_GDP", "Exports_of_goods_and_services__of_GDP",
  "Imports_of_goods_and_services__of_GDP", "Foreign_direct_investment_net_inflows_BoP_current_US",
  "Industry_including_construction_value_added__of_GDP", "Agriculture_forestry_and_fishing_value_added__of_GDP",
  "Services_value_added__of_GDP", "Unemployment_total__of_total_labor_force_modeled_ILO_estimate",
  "Income_share_held_by_lowest_20", "Income_share_held_by_highest_20", "Poverty_headcount_ratio_at_215_a_day_2017_PPP__of_population",

@@

# Subset the dataset with the selected variables
mydata_selected <- mydata %>%
  select(all_of(selected_vars))

@@

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)      # For tidy regression results
library(purrr)      # For functional programming
library(kableExtra) # For advanced table formatting

# Ensure all selected variables are numeric
mydata <- mydata %>%
  mutate(across(everything(), as.numeric))

# Function to run multiple regression for a specific category and country
run_regression <- function(data, category_vars, category_name) {
  # Drop variables with all NA values within the category
  category_vars <- category_vars[sapply(data[category_vars], function(x) !all(is.na(x)))]

@@

  # Create a subset of data with the dependent variable and independent variables
  regression_data <- data %>%
    select(Current_health_expenditure_per_capita_current_US, all_of(category_vars)) %>%
    na.omit()  # Remove rows with NA values

  # Check if there are enough rows to run the regression
  if (nrow(regression_data) == 0) {
    return(NULL)
  }

  # Create the regression formula
  formula <- as.formula(paste("Current_health_expenditure_per_capita_current_US ~", 
                              paste(category_vars, collapse = " + ")))

@@

  # Tidy up the regression results and add a "Significant" column
  tidy_results <- tidy(regression_model) %>%
    mutate(
      Category = category_name,
      Significant = ifelse(p.value <= 0.10, "YES", "NO")  # Add significance column
    )

@@

# Function to run regression for all countries and a specific category
run_regression_all_countries <- function(category_vars, category_name) {
  results <- mydata %>%
    group_by(CountryName) %>%
    group_modify(~ {
      regression_results <- run_regression(.x, category_vars, category_name)
      if (!is.null(regression_results)) {
        return(regression_results)
      } else {
        return(tibble())  # Return empty tibble if no results
      }
    })

  return(results)
}

# Run regression for each category and create tables
economic_results <- run_regression_all_countries(variable_categories$Economic, "Economic")
health_results <- run_regression_all_countries(variable_categories$Health, "Health")
social_results <- run_regression_all_countries(variable_categories$Social, "Social")
government_results <- run_regression_all_countries(variable_categories$Government, "Government")
trade_results <- run_regression_all_countries(variable_categories$Trade, "Trade")

# Combine all results into one table
all_results <- bind_rows(
  economic_results,
  health_results,
  social_results,
  government_results,
  trade_results
)

# Ensure the "Significant" column has no NA values
all_results <- all_results %>%
  mutate(Significant = ifelse(is.na(Significant), "NO", Significant))  # Replace NA with "NO"

# Remove the CountryName column
all_results <- all_results %>%
  select(-CountryName)

@@

# Add category separators
all_results <- all_results %>%
  group_by(Category) %>%
  group_modify(~ {
    # Add a separator row before each category
    separator_row <- tibble(
      `Predictor variables` = paste("Independent variables from", .y$Category, "category"),
      estimate = NA, std.error = NA, statistic = NA, p.value = NA, Significant = NA
    )
    bind_rows(separator_row, .x)
  }) %>%
  ungroup()

# Create an advanced table using kableExtra
all_results %>%
  kable("html", caption = "Regression Results by Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "500px") %>%
  row_spec(which(all_results$Significant == "YES"), background = "#D9F2D9") %>%  # Highlight significant rows in green
  row_spec(which(all_results$`Predictor variables` %in% paste("Independent variables from", unique(all_results$Category), "category")), 
           bold = TRUE, background = "#FFA500") %>%  # Highlight separator rows in orange
  column_spec(1, bold = TRUE)  # Bold the "Predictor variables" column
```

>>>>

We will load the necessary libraries.

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)      
library(kableExtra) 
library(webshot)
library(magick)
library(here)
library(modelsummary)
library(plm)
library(lmtest)
library(gt)
library(webshot2)
library(fixest)
library(lubridate)
library(Metrics)

```

Then we will import the dataset

@@

We will now pick the most useful variables from the large pool of 1400+ variables in our dataset, and we will then categorize these variables based on the aspects they cover. For example, if there is avariable named "Annual GDP growth", we will categorize it as an "Economic" variable. Similarly, if there is a variable named "Unemployment rate" or "Total population", we will categorize them as "Social" variable.

But, before that we will import the training data which was obtained after splitting the dataset into 80% train data and 20% test data. We will be performing all our analysis on train data.

```{r}
train <- readRDS(here("data", "processed-data", "train-data.rds"))
test <- readRDS(here("data", "processed-data", "test-data.rds"))
```

```{r}

#defining the dependent variable
dependent_var <- "Current_health_expenditure_per_capita_current_US"

#selecting 50 variables that are likely to predict the dependent variable
selected_vars <- c(
  # Economic Factors
  "GDP_per_capita_current_US", "GNI_per_capita_Atlas_method_current_US", "GDP_growth_annual_",
  "Inflation_GDP_deflator_annual_", "Domestic_credit_provided_by_financial_sector__of_GDP",
  "Gross_capital_formation__of_GDP", "Exports_of_goods_and_services__of_GDP",
  "Imports_of_goods_and_services__of_GDP", "Foreign_direct_investment_net_inflows_BoP_current_US",
  "Industry_including_construction_value_added__of_GDP", "Agriculture_forestry_and_fishing_value_added__of_GDP",
  "Services_value_added__of_GDP", "Unemployment_total__of_total_labor_force_modeled_ILO_estimate",
  "Income_share_held_by_lowest_20", "Income_share_held_by_highest_20", "Poverty_headcount_ratio_at_215_a_day_2017_PPP__of_population",

@@

# Subset the dataset with the selected variables
train_selected <- train %>%
  select(all_of(selected_vars))

@@

Now that we have selected our variables, we will try to look at the relation between our dependent variable "Health Expenditure Per Capita \$USD" with other variables from a specific category.

```{r}

# converting all columns to numeric except CountryName, CountryCode, and SeriesCode
train <- train %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

#saving the train dataset as we will be using it in future as well

saveRDS(train, file = here("data", "processed-data", "train-data.rds"))

# converting all columns to numeric except CountryName, CountryCode, and SeriesCode
test <- test %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

#saving the train dataset as we will be using it in future as well

saveRDS(test, file = here("data", "processed-data", "test-data.rds"))
```

```{r}

# Function to run regression for a specific category of independent variables
run_regression <- function(data, category_vars, category_name) {
  # dropping variables with all NA values within the category
  category_vars <- category_vars[sapply(data[category_vars], function(x) !all(is.na(x)))]

@@

  # creating a subset of data with the dependent variable and independent variables
  regression_data <- data %>%
    select(Current_health_expenditure_per_capita_current_US, all_of(category_vars)) %>%
    na.omit()  # Remove rows with NA values

  # checking if there are enough rows to run the regression
  if (nrow(regression_data) == 0) {
    return(NULL)
  }

  # creating the regression formula
  formula <- as.formula(paste("Current_health_expenditure_per_capita_current_US ~", 
                              paste(category_vars, collapse = " + ")))

@@

  # Tidy up the regression results and add a "Significant" column
  tidy_results <- tidy(regression_model) %>%
    mutate(
      Category = category_name,
      Significant = ifelse(p.value <= 0.05, "YES", "NO")  # Add significance column
    )

@@

# Run regression for each category and create tables
economic_results <- run_regression(train, variable_categories$Economic, "Economic")
health_results <- run_regression(train, variable_categories$Health, "Health")
social_results <- run_regression(train, variable_categories$Social, "Social")
government_results <- run_regression(train, variable_categories$Government, "Government")
trade_results <- run_regression(train, variable_categories$Trade, "Trade")

# combine all results into one table
all_results <- bind_rows(
  economic_results,
  health_results,
  social_results,
  government_results,
  trade_results
)

# ensuring the "Significant" column has no NA values
all_results <- all_results %>%
  mutate(Significant = ifelse(is.na(Significant), "NO", Significant))  # Replace NA with "NO"

@@

# Adding category separators
all_results <- all_results %>%
  group_by(Category) %>%
  group_modify(~ {
    # Add a separator row before each category
    separator_row <- tibble(
      `Predictor variables` = paste("Independent variables from", .y$Category, "category"),
      estimate = NA, std.error = NA, statistic = NA, p.value = NA, Significant = NA
    )
    bind_rows(separator_row, .x)
  }) %>%
  ungroup()

# creating an advanced table using kableExtra
table_html <- all_results %>%
  kable("html", caption = "Regression Results by Category") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "500px") %>%
  row_spec(which(all_results$Significant == "YES"), background = "#D9F2D9") %>%  # Highlight significant rows in green
  row_spec(which(all_results$`Predictor variables` %in% paste("Independent variables from", unique(all_results$Category), "category")), 
           bold = TRUE, background = "#FFA500") %>%  # Highlight separator rows in orange
  column_spec(1, bold = TRUE)  # Bold the "Predictor variables" column

# Display the table
table_html
```

In the above table, we can observe that a lot of factors affect the dependent variable and a lot of them don't. We can also observe that a lot of our variables have missing values making it unavailable for the regression estimation.

Now, we will save our regression result is a HTML table so that we can use to present it in our manuscript.

```{r}
# Save the table as a HTML file
html_file <- here("results", "tables", "regression_results.html")
save_kable(table_html, file = html_file)

```

Now that we have tested some variables in our model across different categories, we will pick the most useful variable and add them into our preliminary model. We will also add other variable that were not present in the above table. Variable selection is a continuous process and the variables selected in the preliminary model may also be dropped later to make space for new variables that may appear as important.

```{r}
# preliminary regression model

model <- lm(Current_health_expenditure_per_capita_current_US ~ 
              GDP_growth_annual_ + 
              Income_share_held_by_lowest_20 + 
              Control_of_Corruption_Estimate + 
              Life_expectancy_at_birth_total_years + 
Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
              Trade__of_GDP+ Foreign_direct_investment_net_inflows_BoP_current_US+ 
  Access_to_electricity_rural__of_rural_population +
  Age_dependency_ratio__of_workingage_population
  , 
            data = train)

modelsummary(
  model,
  output = here("results", "tables", "preliminary_regression_results.png"),
  title = "This table represents the preliminary regresssion model",
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  notes = list(
    "Standard errors are shown in parentheses.",
    "Significance levels: * p < 0.1, ** p < 0.05, *** p < 0.01."
  )
)

```

The variables have a significant impact on the dependent variable as expected. But, as this is a time series dataset, we must also test for autocorrelation. There are many other tests as well which can be performed on time-series models such ADF test for unit root, and complex models for variance modelling such as ARCH and GARCH can also be used. However, we will only perform simple procedures and try to build a simple model.

# Autocorrelation test

Below, we will perform the autocorrelation test.

```{r}

# Perform the Durbin-Watson test
dw_test <- dwtest(model)

# Extract the test statistic and p-value
dw_statistic <- dw_test$statistic
dw_p_value <- dw_test$p.value

# Create a data frame with the results
dw_results <- data.frame(
  Statistic = dw_statistic,
  P_Value = dw_p_value
)

# Generate a table using gt
dw_table <- gt(dw_results) %>%
  tab_header(
    title = "Durbin-Watson Test Results"
  ) %>%
  fmt_number(
    columns = vars(Statistic, P_Value),
    decimals = 4
  )

# Define the file path for the PNG
png_file <- here::here("results", "figures", "durbin_watson_test_results.png")

# Save the table as a PNG file
gtsave(dw_table, png_file)

```

Fortunately, the Durbin-Watson test is insignificant, and our model does not show presence of autocorrelation.

We can therefore move forward with this model.

Calculating the RMSE of a NULL Model.

```{r}
null_model <- lm(Current_health_expenditure_per_capita_current_US ~ 1, 
                 data = train)

# Calculate residuals
residuals <- residuals(null_model)

# Calculate RMSE manually
null_rmse <- sqrt(mean(residuals^2))

# Display results
summary(null_model)
cat("\nNull Model RMSE:", null_rmse, "\n")
```

 ADF test (combined)
```{r}
library(plm)

# 1. Prepare panel data (handle NAs first)
train_clean <- na.omit(train[, c("CountryName", "Year", "Current_health_expenditure_per_capita_current_US")])

# 2. Check time dimensions
time_counts <- table(train_clean$country)
print(time_counts)

# 3. Run Fisher-type test with reduced lags
pdata <- pdata.frame(train_clean, index = c("CountryName", "Year"))
fisher_test <- purtest(pdata$Current_health_expenditure_per_capita_current_US,
                      test = "madwu",
                      exo = "intercept",
                      lags = 1) # Force shorter lags

```

```{r}
library(kableExtra)
library(here)

# 1. Create properly formatted results table
unit_root_table <- data.frame(
  Test = "Fisher-Type Panel Unit Root Test (ADF)",
  Statistic = fisher_test$statistic$statistic,
  DF = fisher_test$statistic$parameter,
  `p-value` = fisher_test$statistic$p.value,
  check.names = FALSE
)

# 2. Create and save HTML table
unit_root_table %>%
  kbl(format = "html", digits = 4, align = "c",
      caption = "Panel Unit Root Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  add_footnote("Null hypothesis: All series contain unit roots", 
              notation = "none") %>%
  save_kable(here("results", "tables", "unit_root_results.html"))
```

ADF test (Individual countries)

```{r}
library(tseries)

# Option 1: Simple version with try()
adf_results <- by(train, train$CountryName, function(x) {
  try(adf.test(na.omit(x$Current_health_expenditure_per_capita_current_US)),
  silent = TRUE)
})

# Option 2: More robust version with error handling
adf_results <- lapply(split(train, train$CountryName), function(df) {
  if (sum(!is.na(df$Current_health_expenditure_per_capita_current_US)) >= 3) {
    tryCatch({
      adf.test(na.omit(df$Current_health_expenditure_per_capita_current_US))
    }, error = function(e) {
      return(list(statistic = NA, p.value = NA, 
                 note = paste("Error:", e$message)))
    })
  } else {
    return(list(statistic = NA, p.value = NA, 
               note = "Not enough observations"))
  }
})

# Create a clean results table
results_table <- do.call(rbind, lapply(names(adf_results), function(CountryName) {
  res <- adf_results[[CountryName]]
  data.frame(
    Country = CountryName,
    ADF_Statistic = ifelse(is.null(res$statistic), NA, res$statistic),
    P_Value = ifelse(is.null(res$p.value), NA, res$p.value),
    Note = ifelse(is.null(res$note), "", res$note)
  )
}))

print(results_table)
```

### 2025-04-03 code/eda-code/eda.qmd
part #4

@@

# Filtering data for the year 2010
data_2010 <- mydata %>%
  filter(Year == 2010)

@@

```{r}
# Converting column to numeric before plotting
mydata <- mydata %>%
  mutate(Current_health_expenditure__of_GDP = as.numeric(Current_health_expenditure__of_GDP))

# Filtering top 5 countries based on 2010 health expenditure
top_5_countries <- mydata %>%
  filter(Year == 2010) %>%
  arrange(desc(Current_health_expenditure__of_GDP)) %>%
  slice_head(n = 5) %>%
  pull(CountryName)  # Extracting country names

# Filtering data for only 5 countries over time
data_filtered <- mydata %>%
  filter(CountryName %in% top_5_countries)

@@

# Converting columns to numeric 
mydata <- mydata %>%
  mutate(
    Current_health_expenditure__of_GDP = as.numeric(Current_health_expenditure__of_GDP),
    GDP_per_capita_current_US = as.numeric(GDP_per_capita_current_US)
  )

# Filtering data for the year 2010
data_2010 <- mydata %>%
  filter(Year == 2010)

@@

# Converting columns to numeric
mydata <- mydata %>%
  mutate(
    Current_health_expenditure__of_GDP = as.numeric(Current_health_expenditure__of_GDP),
    Hospital_beds_per_1000_people = as.numeric(Hospital_beds_per_1000_people)
  )

# Filtering data for the year 2010
data_2010 <- mydata %>%
  filter(Year == 2010)

@@

# Filtering data for Australia
australia_data <- mydata %>%
  filter(CountryName == "Australia")

@@

# Filter data for France
france_data <- mydata %>%
  filter(CountryName == "France")

@@

@@

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

>>>>

```{r}
# sorting the dataset by time
mydata <- mydata %>% arrange(Year)  # sorting by year

#splitting at 80% of the observations
split_index <- floor(0.8 * nrow(mydata))
train <- mydata[1:split_index, ]
test <- mydata[(split_index + 1):nrow(mydata), ]
```

@@

# Filtering data for the year 2010
data_2010 <- train %>%
  filter(Year == 2010)

@@

```{r}
# Converting column to numeric before plotting
train <- train %>%
  mutate(Current_health_expenditure__of_GDP = as.numeric(Current_health_expenditure__of_GDP))

# Filtering top 5 countries based on 2010 health expenditure
top_5_countries <- train %>%
  filter(Year == 2010) %>%
  arrange(desc(Current_health_expenditure__of_GDP)) %>%
  slice_head(n = 5) %>%
  pull(CountryName)  # Extracting country names

# Filtering data for only 5 countries over time
data_filtered <- train %>%
  filter(CountryName %in% top_5_countries)

@@

# Converting columns to numeric 
train <- train %>%
  mutate(
    Current_health_expenditure__of_GDP = as.numeric(Current_health_expenditure__of_GDP),
    GDP_per_capita_current_US = as.numeric(GDP_per_capita_current_US)
  )

# Filtering data for the year 2010
data_2010 <- train %>%
  filter(Year == 2010)

@@

# Converting columns to numeric
train <- train %>%
  mutate(
    Current_health_expenditure__of_GDP = as.numeric(Current_health_expenditure__of_GDP),
    Hospital_beds_per_1000_people = as.numeric(Hospital_beds_per_1000_people)
  )

# Filtering data for the year 2010
data_2010 <- train %>%
  filter(Year == 2010)

@@

# Filtering data for Australia
australia_data <- train %>%
  filter(CountryName == "Australia")

@@

# Filter data for France
france_data <- train %>%
  filter(CountryName == "France")

@@

```{r}
# Converting column to numeric before plotting
train <- train %>%
  mutate(Current_health_expenditure_per_capita_current_US = as.numeric(Current_health_expenditure_per_capita_current_US))

# Filtering top 5 countries based on 2010 per capita health expenditure
top_5_countries <- train %>%
  filter(Year == 2010) %>%
  arrange(desc(Current_health_expenditure_per_capita_current_US)) %>%
  slice_head(n = 5) %>%
  pull(CountryName)  # Extracting country names

# Filtering data for only 5 countries over time
data_filtered <- train %>%
  filter(CountryName %in% top_5_countries)

y_breaks <- seq(min(data_filtered$Current_health_expenditure_per_capita_current_US, na.rm = TRUE), 
                max(data_filtered$Current_health_expenditure_per_capita_current_US, na.rm = TRUE), 
                length.out = 4)

# Creating the line plot
ggplot(data_filtered, aes(x = Year, y = Current_health_expenditure_per_capita_current_US, color = CountryName, group = CountryName)) +
  geom_line(size = 1) +
  geom_point(size = 2) +  
  labs(
    title = "Current Per Capita Health Expenditure (2010–2020): Top 5 nations",
    x = "Year",
    y = "Health Expenditure (% of GDP)",
    color = "Country"
  ) +
  scale_y_continuous(breaks = y_breaks) +  
  theme_minimal()

ggsave(
  filename = here("results","figures", "per_capita_health_expenditure_plot.png"),  
  plot = last_plot(),  # Saves the last plotted figure
  width = 8, 
  height = 6, 
  dpi = 300
)

@@
